---
title: 'ANTsX: R and Python Introduction'
author: "Avants, Tustison"
date: "5/4/2019"
output: html_document
---


## A normal R code chunk

```{r}
library(reticulate)
matplotlib <- import("matplotlib")
matplotlib$use("Agg", force = TRUE)
x = 42
print(x)
```

## Modify an R variable

In the following chunk, the value of `x` on the right hand side
is `r x`, which was defined in the previous chunk.

```{r}
x = x + 12
print(x)
```

## A Python chunk

This works fine and as expected.

```{python}
x = 42 * 2
print(x)
```

The value of `x` in the Python session is `r py$x`.
It is not the same `x` as the one in R.

## Modify a Python variable

```{python}
x = x + 18
print(x)
```

Retrieve the value of `x` from the Python session again:

```{r}
py$x
```

Assign to a variable in the Python session from R:

```{r}
py$y = 1:5
```

See the value of `y` in the Python session:

```{python}
print(y)
```

## ANTsPy

```{python}
import ants
imgfn = ants.get_data( "r16" )
img = ants.image_read( imgfn )
ants.plot( img )
```

## ANTsR

```{R}
library( ANTsR )
imgfn = getANTsRData( "r16" )
img = antsImageRead( imgfn )
plot( img )
```
